---
title: "Calculations for abiotic methane oxidation"
author: "Lennon et al."
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## Set working directory
```{r}
rm(list=ls())
getwd()
#setwd("~/GitHub/radiolytic")
```

## Energy generated from radon decay
First step in the radon decay chain is the $\alpha$-decay of a single ^222^Rn atom to ^218^Po. 
From Ruzer (2004), Table 14.1, page 365 in "Aerosols Handbook: Measurement, Dosimetry, and Health Effects" editors L.S. Ruzer and N.H. Harley. CRC Press.
```{r}
# Megaelectron-volts (MeV) from Rn decay
Rn.decay <- 5.486 

# Conversion factro from MeV to kilojoules (kJ)
MeV.kJ <- 1.602 * 10^-16 

# kJ of energy in 222Rn to 218Po decay
print (MeV.kJ * Rn.decay) 
```

## Energy required to generate a methyl radical from CH~4~
```{r}
# Bond energy (kJ/mol) to break C-H in CH4 (Smith and March 2007)
bond <- 416

# Avogadro's number
A <- 6.023 * 10^23

# Energy needed to generate a methyl radical
 print(methyl.rad <- bond/A)
```

## How many CH~4~ molecules are there in 2 ppm CH~4~ of air?
```{r}
# assumed ppm of CH4 in air
CH4.air.ppm <- 2

# assumed conctrnation of CH4 in air
CH4.air.conc <- CH4.air.ppm/10^6

# Number of molecules in 1 mole of air
mol.air <- 6.023 * 10^23 

# Number of molecules of CH4 in 1 mole of air at 2 ppm CH4
CH4.molec.mol.air <- mol.air * CH4.air.conc 

# Number of liters in a mole
L.mole <- 22.4

# Liters in a cubic meter (m^3)
L.cm3 <- 1000

# Molecules of CH4 in 1 m^3 of air at 2 ppm CH4
CH4.molec.m3.air <- (CH4.molec.mol.air / L.mole) * L.cm3

print(CH4.molec.m3.air)
```

## How many CH~4~ molecules can be oxidized via direct contact with $\alpha$ particles?
```{r}
# Reported 222Rn concentration (Bq m^-3) in Castanar Cave; decays per second
Rn <- 35000 

# Time to oxidize all CH4 molecules in 2 ppm CH4
CH4.time.secs <- CH4.molec.m3.air/Rn

CH4.time.mins <- CH4.time.secs/60

CH4.time.hrs <- CH4.time.mins/60

CH4.time.days <- CH4.time.hrs/24

CH4.time.yrs <- CH4.time.days/365

print(format(CH4.time.yrs, scientific = TRUE))
```

## How many CH~4~ molecules can be oxidized via ions and radicals generated via $\alpha$ particles?
```{r}

# Ions generated assuming 1 alpha decay generates 10^7 total ions
ions <- 10^7

# Time to oxidize all CH4 molecules in 2 ppm CH4
CH4.time.ions.sec <- CH4.molec.m3.air/(Rn*ions)
  
CH4.time.ions.min <- CH4.time.ions.sec/60

CH4.time.ions.hrs <- CH4.time.ions.min/60

CH4.time.ions.day <- CH4.time.ions.hrs/24

CH4.time.ions.yrs <- CH4.time.ions.day/365

print(format(CH4.time.ions.yrs, scientific = FALSE))
```
